Circumnuclear Media and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-02-12 ms.tex
githole2

Assuming that $v_{w}^{\rm Ia} \gg \sigma$ then by substituting $v_{w}^{\rm Ia}$ (eq.~\ref{eq:vIa}) into equation (\ref{eq:rs_simple}) for the stagnation radius, we find that
\begin{eqnarray}
\left.\frac{r_{\rm Ia}}{r_{\rm s}}\right|_{\rm v_{w}^{\rm Ia}} &\approx& 12.4\eta_{0.02} M_{\bullet,8}^{-1.1}(\epsilon_{\rm Ia}/0.4)^{-1}(\tau_{\star}/t_{\rm h})^{1.68} \nonumber \\
&\approx& 12.4M_{\bullet,8}^{-1.1}(\epsilon_{\rm Ia}/0.4)^{-1}(\tau_{\star}/t_{\rm h})^{0.38}
\label{eq:rIars}
\end{eqnarray}
If $r_{\rm Ia} \gg r_{\rm s}$, then Ia heating can only be approximated as a steady heating source near the stagnation radius for extremely massive SMBHs with $M_{\bullet} \gtrsim 10^9 \Msun$ or in the case of a very young stellar population ($\tau_{\star} \ll t_{\rm h}$).  On the other hand, equation (\ref{eq:rIars}) may underestimate the importance of Ia heating because $r_{\rm Ia}/r_{\rm s}$ will be substantially smaller than its value estimated using $v_{w}^{\rm Ia}$ if the heating rate without SN Ia is less than this.

Even if SN Ia are rare near the stagnation radius, they may cap the SMBH accretion rate by periodically blowing gas out of the nucleus of low mass galaxies.  Between subsequent supernovae, stars release a gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$ interior to the Ia radius, which is gravitationally bound to the SMBH by an energy $E_{\rm bind} \sim M_{\rm g}\sigma^{2}$.  Thus, according to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{w}^{\rm Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence, SN Ia are in principle capable of dynamically clearing out gas from radii $\sim r_{\rm Ia} \gtrsim r_{\rm s}$, especially for low mass BHs with $\sigma \ll v^{\rm Ia}_{w}$.  Thus, even in cases when $r_{\rm s} > r_{\rm Ia}$, the accretion rate is limited to the value, 
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{R_{\rm Ia}}{R_{\rm soi}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note that $\dot{M}_{\rm max,Ia}$ usually exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm max,TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in low mass galaxies.  Also note that this maximum Ia accretion rate only applies if the gas within the CNM is supplied by local stellar mass loss, since the quantity of gaseous mass can be much greater if is fed externally, e.g. due to a galactic merger.

\subsection{Millisecond Pulsars}
 Energy injection from millisecond pulsars (MSPs) is a potentially
important heating source.  If the number of MSPs per unit stellar mass
is $n_{\rm msp}$ and each contributes on average a spin-down
luminosity $\bar{L}_{\rm sd}$, then the resulting heating per unit
volume $e \approx \bar{L}_{\rm sd}n_{\rm msp}\epsilon_{\rm msp}$ results in an
effective heating rate (eq.~ [\ref{eq:vw_eff}]) of 
\begin{eqnarray} v_{w}^{\rm MSP} \approx
30\left(\frac{\epsilon_{\rm msp}}{0.1}\right)^{1/2}\left(\frac{\bar{L}_{\rm
sd}}{10^{34}\,{\rm erg\,s^{-1}}}\right)^{1/2} \eta_{0.02}^{-1/2}\,{\rm
km\,s^{-1}},
 \label{eq:vmsp}
  \end{eqnarray} 
where $\epsilon_{\rm msp}$ is the thermalization efficiency of
the wind, normalized to a value $\epsilon \lesssim 0.1$ inferred based
on modeling the insterstellar mediums of globular clusters
\citep{NaimanSoares-Furtado+:2013a}.  The above estimate adopts a
pulsar density $n_{\rm msp}\simeq3 \times 10^{-40} $ MSPs g$^{-1}$
motivated by the $\sim 30,000$ estimated MSPs in the Milky Way of
stellar mass $6\times 10^{10}M_{\odot}$ (\citealt{Lorimer13}).

The average spin-down luminosity of millisecond pulsars in the field
is estimated to be $\bar{L}_{\rm sd} \sim 10^{34}$ erg s$^{-1}$ based
on the ATNF radio pulsar catalog (\citealt{Manchester+05}), in which
case $v_{w}^{\rm MSP} \lesssim 30$ km s$^{-1}$.  For higher spin-down
luminosities, $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic
of some Fermi-detected pulsars, then the higher value of $v_{w}^{\rm MSP}
\lesssim 300$ km s$^{-1}$ makes MSP heating in principle important
under the optimistc assumption $\epsilon_{\rm msp} = 1$.  On the other hand, the
binaries giving rise to MSPs may be disocciated by stellar
interactions in the dense nuclear cluster, reducing their numbers as
compared to the field population estimate above.  MSPs are thus unlikely to be an important source of heating in comparison to stellar heating.

>>>>

Assuming that $v_{w}^{\rm Ia} \gg \sigma$ then by substituting $v_{w}^{\rm Ia}$ (eq.~\ref{eq:vIa}) into equation (\ref{eq:rs_simple}) for the stagnation radius, we find that
\begin{eqnarray}
\left.\frac{r_{\rm Ia}}{r_{\rm s}}\right|_{\rm v_{w}^{\rm Ia}} &\approx& 12.4\eta_{0.02} M_{\bullet,8}^{-1.1}(\epsilon_{\rm Ia}/0.4)^{-1}(\tau_{\star}/t_{\rm h})^{1.68} \nonumber \\
&\approx& 12.4M_{\bullet,8}^{-1.1}(\epsilon_{\rm Ia}/0.4)^{-1}(\tau_{\star}/t_{\rm h})^{0.38}
\label{eq:rIars}
\end{eqnarray}
If $r_{\rm Ia} \gg r_{\rm s}$, then Ia heating can only be approximated as a steady heating source near the stagnation radius for extremely massive SMBHs with $M_{\bullet} \gtrsim 10^9 \Msun$ or in the case of a very young stellar population ($\tau_{\star} \ll t_{\rm h}$).  On the other hand, equation (\ref{eq:rIars}) may underestimate the importance of Ia heating because $r_{\rm Ia}/r_{\rm s}$ will be substantially smaller than its value estimated using $v_{w}^{\rm Ia}$ if the heating rate excluding SN Ia is less than this.

Even if SN Ia are rare near the stagnation radius, they may cap the SMBH accretion rate by periodically blowing gas out of the nucleus of low mass galaxies.  Between subsequent supernovae, stars release a gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$ interior to the Ia radius, which is gravitationally bound to the SMBH by an energy $E_{\rm bind} \sim M_{\rm g}\sigma^{2}$.  Thus, according to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{w}^{\rm Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence, SN Ia are in principle capable of dynamically clearing out gas from radii $\sim r_{\rm Ia} \gtrsim r_{\rm s}$, at least for low mass BHs with $\sigma \ll v^{\rm Ia}_{w}$.  Thus, even in cases when heating is sufficiently weak that the stagnation radius formally exceeds $r_{\rm Ia}$, the accretion rate is limited to a maximum value, 
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{R_{\rm Ia}}{R_{\rm soi}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note that $\dot{M}_{\rm Ia}$ usually exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in the nuclei of low mass galaxies.  Also note that Ia supernovae only cap the accretion rate to this maximum value if the gas feeding the black hole is supplied by local stellar mass loss: the gaseous mass can become much greater if the nucleus is fed externally, e.g. due to a galaxy merger or to cooling flows developing within the galaxy over much longer timescales (e.g.~\citealt{Ciotti&Ostriker07}).

\subsection{Millisecond Pulsars}
 Energy injection from the spin-down of millisecond pulsars (MSPs) is a potentially
important heating source.  If the number of MSPs per unit stellar mass
is $n_{\rm msp}$ and each contributes on average a spin-down
luminosity $\bar{L}_{\rm sd}$, then the resulting heating per unit
volume $e \approx \bar{L}_{\rm sd}n_{\rm msp}\epsilon_{\rm msp}$ results in an
effective heating rate (eq.~ [\ref{eq:vw_eff}]) of 
\begin{eqnarray} v_{w}^{\rm MSP} \sim
30\left(\frac{\epsilon_{\rm msp}}{0.1}\right)^{1/2}\left(\frac{\bar{L}_{\rm
sd}}{10^{34}\,{\rm erg\,s^{-1}}}\right)^{1/2} \eta_{0.02}^{-1/2}\,{\rm
km\,s^{-1}},
 \label{eq:vmsp}
  \end{eqnarray} 
where $\epsilon_{\rm msp}$ is the thermalization efficiency of
the wind, normalized to a value $\lesssim 0.1$ based on that inferred by modeling the interstellar media of globular clusters
\citep{NaimanSoares-Furtado+:2013a}.  Our numerical estimate assumes a
pulsar density $n_{\rm msp} \sim 3 \times 10^{-40} $ MSPs g$^{-1}$, calculated from the estimated $\sim 30,000$ MSPs in the Milky Way (\citealt{Lorimer13}) of stellar mass $\approx 6\times 10^{10}M_{\odot}$.

The average spin-down luminosity of millisecond pulsars in the field
is estimated to be $\bar{L}_{\rm sd} \sim 10^{34}$ erg s$^{-1}$ based
on the ATNF radio pulsar catalog (\citealt{Manchester+05}), in which
case $v_{w}^{\rm MSP} \lesssim 30$ km s$^{-1}$.  For higher spin-down
luminosities, $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic
of some Fermi-detected pulsars, then the higher value of $v_{w}^{\rm MSP}
\lesssim 300$ km s$^{-1}$ makes MSP heating in principle important
under the optimistc assumption $\epsilon_{\rm msp} = 1$.  However, the
stellar binaries giving rise to MSPs may be disassociated by stellar interactions in the dense nuclear cluster, reducing the numbers of MSPs as compared to the field population estimate above.  MSPs are thus unlikely to be an important source of heating.
